\textbf{Differences in COVID-19 vaccination in the province of Ontario across Health Regions and socio-economic strata}

### 2023-04-24 manuscript_v4/main.tex
updating v4

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi

@@

The COVID-19 pandemic continues to be a worldwide public health concern.
Although vaccines against this disease were rapidly developed,
vaccination uptake has not ben equal across all the segments of the
population. In particular, it has been shown that there have been
differences in vaccine uptake across different segments of the
population. However, there are also differences in vaccination across
geographical areas, which might be important to consider in the
development of future public health policies against COVID-19. In this
study, we examined the relationship between vaccination status (having
received the first dose of a COVID-19 vaccine), and different
socio-economic and geographical factors. Our results show differences in
vaccination due to race/ethnicity, income, Health Regions (geographical
areas used for health service access in Ontario), and their
interactions. In particular, we show that individuals who identified as
Arab/Middle Eastern, Black, or Latin American, had significantly lower
odds of vaccination than White/Caucasian individuals (ORs and 95\%
CIs=0.31{[}0.14,0.68{]}, 0.32{[}0.17,0.60{]}, 0.27{[}0.11,0.66{]},
respectively), and that individuals with a household income below CAD
25,000 who identified as Arab/Middle Eastern (OR=3.08, 95\%
CI={[}1.27,7.47{]}), Black (OR=3.15,95\% CI={[}1.43,6.92{]}), Latin
American (OR=2.81, 95\% CI={[}1. 04,7.59{]}), or that belonged to other
minority groups (OR=4.63,95\% CI={[}2.34,9.13{]}) had higher odds of
vaccination than individuals from the same racial/ethnic group in higher
income brackets. Finally, we also identified lower odds of vaccination
within certain minority groups in the West Health Region. This study
shows that there is an ongoing need to better understand and address
differences in vaccination uptake across diverse segments of the
population that have been largely impacted by the pandemic.

@@

This problematic in the rollout of the COVID-19 vaccines is a
multifaceted issue resulting from, among other things, the development
of new variants due to inadequate public health
measures\textsuperscript{\protect\hyperlink{ref-li2021}{7}}, inequality
in vaccine access between high- and low-income
countries\textsuperscript{\protect\hyperlink{ref-gerretsen2021}{8},\protect\hyperlink{ref-tamey2022}{9}},
vaccine
hesitancy\textsuperscript{\protect\hyperlink{ref-nafilyan2021}{10}}, and
differences in vaccination uptake across different segments of the
population\textsuperscript{\protect\hyperlink{ref-malik2020}{11}}. In
particular, it is well established that differences in vaccination
uptake have been present even in countries that have had ample access to
vaccines since 2020 (such as the US, the UK, and Canada), where lower
vaccine uptake has been observed within racial minorities (i.e.,
individuals that identify as Black, Asian, or Indigenous), and in
individuals within low income
brackets\textsuperscript{\protect\hyperlink{ref-willis2021}{12}--\protect\hyperlink{ref-khubchandani2021}{15}}.
Reasons given for lower vaccine uptake in these cases have included
medical mistrust due to systemic medical
racism\textsuperscript{\protect\hyperlink{ref-stoler2021}{14}}, mistrust
in vaccines\textsuperscript{\protect\hyperlink{ref-willis2021}{12}}, and
the influence of conspiracy
theories\textsuperscript{\protect\hyperlink{ref-bogart2021}{16}--\protect\hyperlink{ref-freeman2020}{18}}.
Moreover, in the case of Canada, lower vaccine uptake has been observed
in young individuals, those with a low educational level, households
with children, those without a regular healthcare provider, individuals
that identify as part of a visible minorities or Indigenous, and those
that are financially
unstable\textsuperscript{\protect\hyperlink{ref-guay2022}{19}--\protect\hyperlink{ref-hussain2022}{21}}.

@@

This need is particularly important in the case of Ontario, the most
populated province in Canada. Between 2006 and 2019, Ontario provided
healthcare access to its inhabitants using 14 intra-provincial divisions
called the Local Health Integrated Networks (LHINs). However, this
approach was complex, bureaucratic, and led to systemic
inequalities\textsuperscript{\protect\hyperlink{ref-tsasis2012}{30}}. In
late 2019, the 14 LHINs were phased out and the areas they covered were
incorporated into 6 Health Regions (North East, North West, Central,
Toronto, West, and East) in an effort to improve the healthcare system
of the province\textsuperscript{\protect\hyperlink{ref-dong2022}{31}}.
Because the adoption of the Health Regions is relatively recent there is
an ongoing need to analyze the impact of this measure and identify the
existence of intra-regional differences that might exist, and which
could be specially important in the context of the COVID-19 pandemic.

In this study, we analyzed differences in self-reported COVID-19
vaccination status in Ontario using socio-economic (e.g., income,
racial/ethnic identification), temporal (through a consecutive period of
three months in late 2021), and spatial information (using the Health
Regions of Ontario) to identify the existence of differences that might
need to be addressed to ensure that the healthcare system of the
province is more inclusive and that responds to the needs of its most
vulnerable population.

@@

We used data from the \emph{Survey of COVID-19 related Behaviours and
Attitudes}, a repeated cross sectional survey focused on the Canadian
province of Ontario that was commissioned by the Fields Institute for
Research in Mathematical Sciences (henceforth Fields) and the
Mathematical Modelling of COVID-19 Task Force under ethical guidance
from the University of Toronto, and which ran between September 30th,
2021 and January 17th,2022. The survey collected socio-economic
information from participants (Table~\ref{tbl-covariates}), recorded
their location (using the nearest municipality), the date of access to
the survey, and asked information on vaccination status by using the
question ``Have you received the first dose of the COVID vaccine?'',
with possible answers ``yes'' and ``no''. The original dataset contained
39,029 entries (where each entry corresponded to a unique respondent).

@@

\hypertarget{tbl-covariates}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.3014}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.6986}}@{}}
\caption{\label{tbl-covariates}Socio-economic factors from the Fields
COVID-19 survey}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Levels
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Levels
\end{minipage} \\
\midrule()
\endhead
Age group & 16-34,35-54,55 and over \\
Income bracket (CAD) & under 25,000, 25,000-59,999, 60,000 and above \\
Race/ethnicity & Arab/Middle Eastern, Black, East Asian/Pacific
Islander, Indigenous, Latin American, Mixed, South Asian, White
Caucasian, Other \\
\bottomrule()
\end{longtable}

@@

We used a logistic regression model to estimate the probability of
vaccination depending on the socio-economic factors described in
Table~\ref{tbl-covariates}, the month when the survey was answered, the
Health Regions from Ontario indicated in Section~\ref{sec-data}, and the
interactions between Race and Health Region, and Race and income, as
previous studies have shown that socio-economic factors and their
interactions are significant predictors of intent of vaccination and
vaccination
status\textsuperscript{\protect\hyperlink{ref-nguyen2022}{32}--\protect\hyperlink{ref-cnat2022a}{34}}.

The model was fitted first to the clean dataset to obtain uncorrected
estimates. Because we identified differences between the proportions of
all the socio-economic factors included in the analysis
(Table~\ref{tbl-covariates}) and the Census data for Ontario, we used an
iterative proportional fitting procedure
(\emph{raking})\textsuperscript{\protect\hyperlink{ref-deming1940}{35}}
to correct the data using Census socio-economic data and Health Region
population totals, in order to obtain corrected estimates from the
model. Details regarding the correction can be found in the Appendix.
All analyses were conducted in R 4.2.2 using the packages
\texttt{survey}\textsuperscript{\protect\hyperlink{ref-lumley2011}{36}},\texttt{tidyverse}\textsuperscript{\protect\hyperlink{ref-wickham2019}{37}},
and \texttt{quarto}\textsuperscript{\protect\hyperlink{ref-quarto}{38}}.

@@

\begin{tabular}{lcc}
\toprule
\textbf{Variable} & \textbf{no}, N = 958 & \textbf{yes}, N = 2,591\\
\midrule
\textbf{Income (CAD)} &  & \\
\hspace{1em}60000 and above & 305 (23\%) & 1,048 (77\%)\\
\hspace{1em}25000-59999 & 253 (28\%) & 636 (72\%)\\
\hspace{1em}under 25000 & 400 (31\%) & 907 (69\%)\\
\textbf{Age Group} &  & \\
\hspace{1em}16-34 & 409 (27\%) & 1,111 (73\%)\\
\hspace{1em}35-54 & 252 (26\%) & 712 (74\%)\\
\hspace{1em}55 and over & 297 (28\%) & 768 (72\%)\\
\textbf{Health Region} &  & \\
\hspace{1em}Toronto & 371 (28\%) & 952 (72\%)\\
\hspace{1em}Central & 224 (28\%) & 580 (72\%)\\
\hspace{1em}East & 135 (23\%) & 448 (77\%)\\
\hspace{1em}West & 228 (27\%) & 611 (73\%)\\
\textbf{Month} &  & \\
\hspace{1em}October & 469 (27\%) & 1,263 (73\%)\\
\hspace{1em}November & 376 (28\%) & 980 (72\%)\\
\hspace{1em}December & 113 (25\%) & 348 (75\%)\\
\textbf{Race} &  & \\
\hspace{1em}White/Caucasian & 233 (18\%) & 1,079 (82\%)\\
\hspace{1em}Arab/Middle Eastern & 76 (36\%) & 138 (64\%)\\
\hspace{1em}Black & 114 (38\%) & 184 (62\%)\\
\hspace{1em}East Asian/Pacific Islander & 69 (23\%) & 234 (77\%)\\
\hspace{1em}Indigenous & 76 (40\%) & 115 (60\%)\\
\hspace{1em}Latin American & 69 (38\%) & 111 (62\%)\\
\hspace{1em}Mixed & 105 (34\%) & 205 (66\%)\\
\hspace{1em}Other & 128 (35\%) & 239 (65\%)\\
\hspace{1em}South Asian & 88 (24\%) & 286 (76\%)\\
\bottomrule
\multicolumn{3}{l}{\rule{0pt}{1em}\textsuperscript{1} n (\%)}\\
\end{tabular}
\endgroup{}
\end{table}

@@

Table~\ref{tbl-model} shows the results of the logistic regression
models (for the uncorrected and corrected data) on vaccination status
using socio-economic factors (age group, income, race), months covered
by the survey (October, November, and December), geographical areas
(Health Regions) and the interactions between income and race and Health
Region and race. The corrected results show that there were no
statistically significant differences in vaccination rate between the
age groups or across the months covered by the survey as in both cases
the odd ratios were similar. However, between the different household
income brackets, individuals with an income under CAD 25,000 or between
CAD 25,000-59,999 had significantly lower odds of vaccination than those
with an income above CAD 60,000 (ORs=0.37 and 0.59,
\emph{p}\textless0.001 and \emph{p}=0.010, respectively). Within
Race/Ethnicity, individuals who identified as Arab/Middle Eastern,
Black, or Latin American, had significantly lower odds of vaccination
than those in the White/Caucasian group (ORs=0.32, 0.32, 0.27, and
\emph{p}=0.003, \emph{p}\textless0.001 and \emph{p}=0.004,
respectively); additionally, those individuals that reported to belong
to the ``Other'' Race/Ethnicity group (which included the Southeast
Asian, Filipino, West Asian, and Minorities Not Identified Elsewhere
groups according to the Census) had even lower odds of vaccination than
the other minority groups (OR=0.22, \emph{p}\textless0.001). Regarding
Health Regions, individuals that reported living in the West Health
Region (which comprises the regions of Waterloo and Niagara, the
counties of Wellington, Essex, and Lambton, and the cities of Hamilton,
Haldimand, Brant, and Chatham-Kent) had significantly higher odds of
vaccination than those in the Health Region of Toronto (OR=1.54,
\emph{p}=0.031).

@@

\footnotesize
\renewcommand{\arraystretch}{0.5}

\hypertarget{tbl-model}{}
\begin{longtable}{lcccccc}
\caption{\label{tbl-model}Multiple Regression Analysis-Predictors of Vaccination Status }\tabularnewline

\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{\textbf{Uncorrected}} & \multicolumn{3}{c}{\textbf{Corrected}} \\
\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7}
\textbf{Characteristic} & \textbf{OR} & \textbf{95\% CI} & \textbf{p-value} & \textbf{OR} & \textbf{95\% CI} & \textbf{p-value}\\
\midrule
\endfirsthead
\multicolumn{7}{@{}l}{\textit{(continued)}}\\
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{\textbf{Uncorrected}} & \multicolumn{3}{c}{\textbf{Corrected}} \\
\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7}
\textbf{Characteristic} & \textbf{OR} & \textbf{95\% CI} & \textbf{p-value} & \textbf{OR} & \textbf{95\% CI} & \textbf{p-value}\\
\midrule
\endhead

\endfoot
\bottomrule
\endlastfoot
\textbf{Age Group} &  &  &  &  &  & \\
\hspace{1em}16-34 & — & — &  & — & — & \\
\hspace{1em}35-54 & 0.93 & 0.77, 1.14 & 0.5 & 0.90 & 0.67, 1.21 & 0.5\\
\hspace{1em}55 and over & 0.74 & 0.61, 0.89 & 0.002 & 0.99 & 0.74, 1.32 & >0.9\\
\textbf{Month} &  &  &  &  &  & \\
\hspace{1em}October & — & — &  & — & — & \\
\hspace{1em}November & 0.95 & 0.81, 1.12 & 0.6 & 0.92 & 0.71, 1.18 & 0.5\\
\hspace{1em}December & 1.16 & 0.91, 1.49 & 0.2 & 1.16 & 0.78, 1.71 & 0.5\\
\textbf{Income (CAD)} &  &  &  &  &  & \\
\hspace{1em}60000 and above & — & — &  & — & — & \\
\hspace{1em}25000-59999 & 0.59 & 0.41, 0.84 & 0.003 & 0.59 & 0.39, 0.88 & 0.010\\
\hspace{1em}under 25000 & 0.41 & 0.29, 0.58 & <0.001 & 0.37 & 0.25, 0.56 & <0.001\\
\textbf{Race} &  &  &  &  &  & \\
\hspace{1em}White/Caucasian & — & — &  & — & — & \\
\hspace{1em}Arab/Middle Eastern & 0.24 & 0.12, 0.49 & <0.001 & 0.31 & 0.14, 0.68 & 0.003\\
\hspace{1em}Black & 0.30 & 0.17, 0.54 & <0.001 & 0.32 & 0.17, 0.60 & <0.001\\
\hspace{1em}East Asian/Pacific Islander & 0.73 & 0.36, 1.51 & 0.4 & 1.15 & 0.50, 2.66 & 0.7\\
\hspace{1em}Indigenous & 0.38 & 0.17, 0.82 & 0.014 & 0.45 & 0.20, 1.04 & 0.061\\
\hspace{1em}Latin American & 0.28 & 0.13, 0.59 & <0.001 & 0.27 & 0.11, 0.66 & 0.004\\
\hspace{1em}Mixed & 0.58 & 0.30, 1.11 & 0.10 & 0.63 & 0.24, 1.62 & 0.3\\
\hspace{1em}Other & 0.20 & 0.11, 0.35 & <0.001 & 0.22 & 0.12, 0.41 & <0.001\\
\hspace{1em}South Asian & 0.79 & 0.44, 1.43 & 0.4 & 0.90 & 0.48, 1.67 & 0.7\\
\textbf{Health Region} &  &  &  &  &  & \\
\hspace{1em}Toronto & — & — &  & — & — & \\
\hspace{1em}Central & 1.30 & 0.84, 2.00 & 0.2 & 1.47 & 0.92, 2.36 & 0.11\\
\hspace{1em}East & 1.54 & 1.01, 2.34 & 0.044 & 1.41 & 0.90, 2.22 & 0.14\\
\hspace{1em}West & 1.35 & 0.94, 1.93 & 0.10 & 1.54 & 1.04, 2.29 & 0.031\\
\textbf{Income (CAD) * Race} &  &  &  &  &  & \\
\hspace{1em}25000-59999 * Arab/Middle Eastern & 2.19 & 0.95, 5.05 & 0.067 & 1.85 & 0.68, 4.99 & 0.2\\
\hspace{1em}under 25000 * Arab/Middle Eastern & 2.95 & 1.39, 6.24 & 0.005 & 3.08 & 1.27, 7.47 & 0.013\\
\hspace{1em}25000-59999 * Black & 1.18 & 0.59, 2.36 & 0.6 & 1.31 & 0.58, 2.98 & 0.5\\
\hspace{1em}under 25000 * Black & 2.84 & 1.46, 5.53 & 0.002 & 3.15 & 1.43, 6.92 & 0.004\\
\hspace{1em}25000-59999 * East Asian/Pacific Islander & 0.97 & 0.44, 2.10 & >0.9 & 0.42 & 0.17, 1.06 & 0.066\\
\hspace{1em}under 25000 * East Asian/Pacific Islander & 1.90 & 0.90, 4.01 & 0.094 & 1.15 & 0.47, 2.84 & 0.8\\
\hspace{1em}25000-59999 * Indigenous & 1.77 & 0.72, 4.32 & 0.2 & 1.33 & 0.46, 3.80 & 0.6\\
\hspace{1em}under 25000 * Indigenous & 1.60 & 0.71, 3.60 & 0.3 & 1.41 & 0.54, 3.72 & 0.5\\
\hspace{1em}25000-59999 * Latin American & 0.89 & 0.38, 2.10 & 0.8 & 1.25 & 0.45, 3.45 & 0.7\\
\hspace{1em}under 25000 * Latin American & 3.10 & 1.34, 7.21 & 0.009 & 2.81 & 1.04, 7.59 & 0.041\\
\hspace{1em}25000-59999 * Mixed & 0.87 & 0.39, 1.93 & 0.7 & 0.87 & 0.33, 2.33 & 0.8\\
\hspace{1em}under 25000 * Mixed & 1.27 & 0.65, 2.49 & 0.5 & 1.12 & 0.38, 3.33 & 0.8\\
\hspace{1em}25000-59999 * Other & 5.50 & 2.43, 12.5 & <0.001 & 6.96 & 2.67, 18.2 & <0.001\\
\hspace{1em}under 25000 * Other & 4.09 & 2.27, 7.37 & <0.001 & 4.63 & 2.34, 9.13 & <0.001\\
\hspace{1em}25000-59999 * South Asian & 1.14 & 0.55, 2.38 & 0.7 & 1.22 & 0.52, 2.87 & 0.7\\
\hspace{1em}under 25000 * South Asian & 1.58 & 0.82, 3.05 & 0.2 & 2.01 & 0.93, 4.34 & 0.075\\
\textbf{Race * Health Region} &  &  &  &  &  & \\
\hspace{1em}Arab/Middle Eastern * Central & 0.76 & 0.33, 1.71 & 0.5 & 0.66 & 0.25, 1.69 & 0.4\\
\hspace{1em}Black * Central & 0.48 & 0.23, 1.02 & 0.055 & 0.44 & 0.19, 0.99 & 0.046\\
\hspace{1em}East Asian/Pacific Islander * Central & 0.84 & 0.38, 1.89 & 0.7 & 0.98 & 0.38, 2.53 & >0.9\\
\hspace{1em}Indigenous * Central & 0.60 & 0.24, 1.50 & 0.3 & 0.63 & 0.22, 1.78 & 0.4\\
\hspace{1em}Latin American * Central & 0.69 & 0.28, 1.71 & 0.4 & 0.66 & 0.22, 1.93 & 0.4\\
\hspace{1em}Mixed * Central & 0.63 & 0.30, 1.35 & 0.2 & 0.73 & 0.24, 2.23 & 0.6\\
\hspace{1em}Other * Central & 0.98 & 0.49, 1.95 & >0.9 & 0.81 & 0.36, 1.81 & 0.6\\
\hspace{1em}South Asian * Central & 0.71 & 0.34, 1.47 & 0.4 & 0.55 & 0.25, 1.21 & 0.14\\
\hspace{1em}Arab/Middle Eastern * East & 0.57 & 0.20, 1.63 & 0.3 & 0.43 & 0.13, 1.43 & 0.2\\
\hspace{1em}Black * East & 0.84 & 0.37, 1.89 & 0.7 & 0.86 & 0.35, 2.11 & 0.7\\
\hspace{1em}East Asian/Pacific Islander * East & 0.82 & 0.30, 2.24 & 0.7 & 0.86 & 0.29, 2.54 & 0.8\\
\hspace{1em}Indigenous * East & 0.56 & 0.21, 1.45 & 0.2 & 0.70 & 0.23, 2.11 & 0.5\\
\hspace{1em}Latin American * East & 0.80 & 0.28, 2.25 & 0.7 & 1.03 & 0.31, 3.40 & >0.9\\
\hspace{1em}Mixed * East & 0.70 & 0.31, 1.60 & 0.4 & 0.92 & 0.28, 3.05 & 0.9\\
\hspace{1em}Other * East & 0.87 & 0.38, 2.02 & 0.7 & 1.08 & 0.40, 2.91 & 0.9\\
\hspace{1em}South Asian * East & 0.50 & 0.20, 1.24 & 0.14 & 0.51 & 0.19, 1.42 & 0.2\\
\hspace{1em}Arab/Middle Eastern * West & 1.18 & 0.49, 2.82 & 0.7 & 1.02 & 0.37, 2.77 & >0.9\\
\hspace{1em}Black * West & 0.78 & 0.36, 1.67 & 0.5 & 0.77 & 0.32, 1.82 & 0.5\\
\hspace{1em}East Asian/Pacific Islander * West & 0.54 & 0.24, 1.21 & 0.14 & 0.52 & 0.20, 1.36 & 0.2\\
\hspace{1em}Indigenous * West & 0.45 & 0.19, 1.05 & 0.064 & 0.40 & 0.14, 1.12 & 0.082\\
\hspace{1em}Latin American * West & 0.99 & 0.39, 2.48 & >0.9 & 0.96 & 0.33, 2.79 & >0.9\\
\hspace{1em}Mixed * West & 0.53 & 0.25, 1.10 & 0.088 & 0.37 & 0.12, 1.15 & 0.085\\
\hspace{1em}Other * West & 0.55 & 0.27, 1.12 & 0.10 & 0.41 & 0.18, 0.92 & 0.032\\
\hspace{1em}South Asian * West & 0.50 & 0.24, 1.08 & 0.078 & 0.41 & 0.18, 0.95 & 0.037\\*
\multicolumn{7}{l}{\rule{0pt}{1em}\textsuperscript{1} OR = Odds Ratio, CI = Confidence Interval}\\
\end{longtable}

\normalsize

@@

Additionally, significant higher vaccination odds were identified in the
West Health Region when compared to the Health Region of Toronto
(Table~\ref{tbl-model}). The West Health Region comprises the regions of
Waterloo and Niagara, the counties of Wellington, Essex and Lambton, and
the cities of Hamilton, Haldimand, Brant, and Chatham-Kent. In this
case, a possible rationale for the results is the fact that in the
survey, about 47\% of the entries for this Health Region corresponded to
White/Caucasian individuals, who reported an overall 83\% vaccination
rate (Supplementary Table A-7). However, the interaction effect of
Health Region and race was also significant in the case of individuals
identifying as South Asian or other minorities not included in the
survey Table~\ref{tbl-model}. In this case, the results of the
interaction term in the model indicate that the odds of vaccination for
those within the South Asian and Other minority groups in the West
Region decreased when compared to the other Health Regions
(Supplementary Figure A-4).

>>>>

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, frame hidden, enhanced, breakable, interior hidden, sharp corners]}{\end{tcolorbox}}\fi

@@

The COVID-19 pandemic continues to be a worldwide public health concern.
Although vaccines against this disease were rapidly developed,
vaccination uptake has not ben equal across all the segments of the
population. In particular, it has been shown that there have been
differences in vaccine uptake across different segments of the
population. However, there are also differences in vaccination across
geographical areas, which might be important to consider in the
development of future public health policies against COVID-19. In this
study, we examined the relationship between vaccination status (having
received the first dose of a COVID-19 vaccine), and different
socio-economic and geographical factors. Our results show that during
the last three months of 2021, individuals in certain equity-deserving
groups (visible minorities) were three times less likely to be
vaccinated than White/Caucasian individuals across the province and that
in some cases, within these groups individuals in low income brackets
had significantly higher odds of vaccination when compared to their
peers in high income brackets. Finally, we identified significantly
lower odds of vaccination in the West Health Region of Ontario within
certain equity-deserving groups. This study shows that there is an
ongoing need to better understand and address differences in vaccination
uptake across diverse segments of the population of Ontario that have
been largely impacted by the pandemic.

@@

This problematic in the rollout of the COVID-19 vaccines is a
multifaceted issue resulting from, among other things, the development
of new variants due to inadequate public health
measures\textsuperscript{\protect\hyperlink{ref-li2021}{7}}, inequality
in vaccine access between high- and low-income
countries\textsuperscript{\protect\hyperlink{ref-gerretsen2021}{8},\protect\hyperlink{ref-tamey2022}{9}},
vaccine
hesitancy\textsuperscript{\protect\hyperlink{ref-nafilyan2021}{10}}, and
differences in vaccination uptake across different segments of the
population\textsuperscript{\protect\hyperlink{ref-malik2020}{11}}. In
particular, it is well established that differences in vaccination
uptake have been present even in countries that have had ample access to
vaccines since 2020 (such as the US, the UK, and Canada), where lower
vaccine uptake has been observed within certain racial groups (i.e.,
individuals that identify as Black, Asian, or Indigenous), and in
individuals within low income
brackets\textsuperscript{\protect\hyperlink{ref-willis2021}{12}--\protect\hyperlink{ref-khubchandani2021}{15}}.
Reasons given for lower vaccine uptake in these cases have included
medical mistrust due to systemic medical
racism\textsuperscript{\protect\hyperlink{ref-stoler2021}{14}}, mistrust
in vaccines\textsuperscript{\protect\hyperlink{ref-willis2021}{12}}, and
the influence of conspiracy
theories\textsuperscript{\protect\hyperlink{ref-bogart2021}{16}--\protect\hyperlink{ref-freeman2020}{18}}.
Moreover, in the case of Canada, lower vaccine uptake has been observed
in young individuals, those with a low educational level, households
with children, those without a regular healthcare provider, individuals
that identify as part of certain equity-deserving groups, and those with
a low household
income\textsuperscript{\protect\hyperlink{ref-guay2022}{19}--\protect\hyperlink{ref-hussain2022}{21}}.

@@

This need is particularly important in the case of Ontario, the most
populated province in Canada. Between 2006 and 2019, Ontario provided
healthcare access to its inhabitants using 14 intra-provincial divisions
called the Local Health Integrated Networks (LHINs). However, this
approach was complex, bureaucratic, and led to systemic
inequalities\textsuperscript{\protect\hyperlink{ref-tsasis2012}{30}}. In
late 2019, the 14 LHINs were phased out and the areas they covered were
incorporated into 6 Health Regions (North East, North West, Central,
Toronto, West, and East) in an effort to improve the healthcare system
of the province\textsuperscript{\protect\hyperlink{ref-dong2022}{31}}.
Because the adoption of the Health Regions is relatively recent, there
is an ongoing need to analyze the impact of this measure and identify
disparities in health access that might exist across the Health Regions,
which can be specially important in the context of the COVID-19
pandemic.

Therefore, in this study we hypothesized that there were differences in
vaccination uptake between the Health Regions of Ontario during the last
quarter of 2021.To better contextualize potential differences, we also
included socio-economic factors in our analysis in order to identify in
which demographic groups these differences were significant.

@@

We used data from the \emph{Survey of COVID-19 related Behaviours and
Attitudes}, a repeated cross sectional survey focused on the Canadian
province of Ontario that was commissioned by the Fields Institute for
Research in Mathematical Sciences (henceforth Fields) and the
Mathematical Modelling of COVID-19 Task Force under ethical guidance
from the University of Toronto, and which ran between September 30th,
2021 and January 17th,2022. The survey collected socio-economic
information from participants (\textbf{?@tbl-covariates}), recorded
their location (using the nearest municipality), the date of access to
the survey, and asked information on vaccination status by using the
question ``Have you received the first dose of the COVID vaccine?'',
with possible answers ``yes'' and ``no''. The original dataset contained
39,029 entries (where each entry corresponded to a unique respondent).

@@

@@

We used a logistic regression model to estimate the probability of
vaccination depending on the socio-economic factors described in
\textbf{?@tbl-covariates}, the month when the survey was answered, the
Health Regions from Ontario indicated in Section~\ref{sec-data}, and the
interactions between Race and Health Region, and Race and income, as
previous studies have shown that socio-economic factors and their
interactions are significant predictors of intent of vaccination and
vaccination
status\textsuperscript{\protect\hyperlink{ref-nguyen2022}{32}--\protect\hyperlink{ref-cnat2022a}{34}}.

The model was fitted first to the clean dataset to obtain uncorrected
estimates. Because we identified differences between the proportions of
all the socio-economic factors included in the analysis
(\textbf{?@tbl-covariates}) and the Census data for Ontario, we used an
iterative proportional fitting procedure
(\emph{raking})\textsuperscript{\protect\hyperlink{ref-deming1940}{35}}
to correct the data using Census socio-economic data and Health Region
population totals, in order to obtain corrected estimates from the
model. Details regarding the correction can be found in the Appendix.
All analyses were conducted in R 4.2.2 using the packages
\texttt{survey}\textsuperscript{\protect\hyperlink{ref-lumley2011}{36}},\texttt{tidyverse}\textsuperscript{\protect\hyperlink{ref-wickham2019}{37}},
and \texttt{quarto}\textsuperscript{\protect\hyperlink{ref-quarto}{38}}.

@@

\begin{tabular}{lccc}
\toprule
\textbf{Variable} & \textbf{no}, N = 958 & \textbf{yes}, N = 2,591 & \textbf{p-value}\\
\midrule
\textbf{Income (CAD)} &  &  & <0.001\\
\hspace{1em}60000 and above & 305 (23\%) & 1,048 (77\%) & \\
\hspace{1em}25000-59999 & 253 (28\%) & 636 (72\%) & \\
\hspace{1em}under 25000 & 400 (31\%) & 907 (69\%) & \\
\textbf{Age Group} &  &  & 0.7\\
\hspace{1em}16-34 & 409 (27\%) & 1,111 (73\%) & \\
\hspace{1em}35-54 & 252 (26\%) & 712 (74\%) & \\
\hspace{1em}55 and over & 297 (28\%) & 768 (72\%) & \\
\textbf{Health Region} &  &  & 0.14\\
\hspace{1em}Toronto & 371 (28\%) & 952 (72\%) & \\
\hspace{1em}Central & 224 (28\%) & 580 (72\%) & \\
\hspace{1em}East & 135 (23\%) & 448 (77\%) & \\
\hspace{1em}West & 228 (27\%) & 611 (73\%) & \\
\textbf{Month} &  &  & 0.4\\
\hspace{1em}October & 469 (27\%) & 1,263 (73\%) & \\
\hspace{1em}November & 376 (28\%) & 980 (72\%) & \\
\hspace{1em}December & 113 (25\%) & 348 (75\%) & \\
\textbf{Race} &  &  & <0.001\\
\hspace{1em}White/Caucasian & 233 (18\%) & 1,079 (82\%) & \\
\hspace{1em}Arab/Middle Eastern & 76 (36\%) & 138 (64\%) & \\
\hspace{1em}Black & 114 (38\%) & 184 (62\%) & \\
\hspace{1em}East Asian/Pacific Islander & 69 (23\%) & 234 (77\%) & \\
\hspace{1em}Indigenous & 76 (40\%) & 115 (60\%) & \\
\hspace{1em}Latin American & 69 (38\%) & 111 (62\%) & \\
\hspace{1em}Mixed & 105 (34\%) & 205 (66\%) & \\
\hspace{1em}Other & 128 (35\%) & 239 (65\%) & \\
\hspace{1em}South Asian & 88 (24\%) & 286 (76\%) & \\
\bottomrule
\multicolumn{4}{l}{\rule{0pt}{1em}\textsuperscript{1} n (\%)}\\
\multicolumn{4}{l}{\rule{0pt}{1em}\textsuperscript{2} Pearson's Chi-squared test}\\
\end{tabular}
\endgroup{}
\end{table}

@@

\textbf{?@tbl-model} shows the results of the logistic regression models
(for the uncorrected and corrected data) on vaccination status using
socio-economic factors (age group, income, race), months covered by the
survey (October, November, and December), geographical areas (Health
Regions) and the interactions between income and race and Health Region
and race. The corrected results show that there were no statistically
significant differences in vaccination rate between the age groups or
across the months covered by the survey as in both cases the odd ratios
were similar. However, between the different household income brackets,
individuals with an income under CAD 25,000 or between CAD 25,000-59,999
had significantly lower odds of vaccination than those with an income
above CAD 60,000 (ORs=0.37 and 0.59, \emph{p}\textless0.001 and
\emph{p}=0.010, respectively). Within Race/Ethnicity, individuals who
identified as Arab/Middle Eastern, Black, or Latin American, had
significantly lower odds of vaccination than those in the
White/Caucasian group (ORs=0.32, 0.32, 0.27, and \emph{p}=0.003,
\emph{p}\textless0.001 and \emph{p}=0.004, respectively); additionally,
those individuals that reported to belong to the ``Other''
Race/Ethnicity group (which included the Southeast Asian, Filipino, West
Asian, and Minorities Not Identified Elsewhere groups according to the
Census) had even lower odds of vaccination than the other minority
groups (OR=0.22, \emph{p}\textless0.001). Regarding Health Regions,
individuals that reported living in the West Health Region (which
comprises the regions of Waterloo and Niagara, the counties of
Wellington, Essex, and Lambton, and the cities of Hamilton, Haldimand,
Brant, and Chatham-Kent) had significantly higher odds of vaccination
than those in the Health Region of Toronto (OR=1.54, \emph{p}=0.031).

@@

\includegraphics{main_files/figure-pdf/fig-model-uncorr-1.pdf} \hfill{}

\caption{\label{fig-model-uncorr}Coefficient estimates and confidence
intervals for the uncorrected model. For the interaction terms, only
those with a statistically significant values are shown.}

\includegraphics{main_files/figure-pdf/fig-model-corr-1.pdf} \hfill{}

\caption{\label{fig-model-corr}Coefficient estimates and confidence
intervals for the corrected model. For the interaction terms, only those
with a statistically significant values are shown.}

@@

Additionally, significant higher vaccination odds were identified in the
West Health Region when compared to the Health Region of Toronto
(\textbf{?@tbl-model}). The West Health Region comprises the regions of
Waterloo and Niagara, the counties of Wellington, Essex and Lambton, and
the cities of Hamilton, Haldimand, Brant, and Chatham-Kent. In this
case, a possible rationale for the results is the fact that in the
survey, about 47\% of the entries for this Health Region corresponded to
White/Caucasian individuals, who reported an overall 83\% vaccination
rate (Supplementary Table A-7). However, the interaction effect of
Health Region and race was also significant in the case of individuals
identifying as South Asian or other minorities not included in the
survey \textbf{?@tbl-model}. In this case, the results of the
interaction term in the model indicate that the odds of vaccination for
those within the South Asian and Other minority groups in the West
Region decreased when compared to the other Health Regions
(Supplementary Figure A-4).
